---
title: "Getting started with CSOA"
author: "Andrei-Florian Stoica"
package: CSOA
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Getting started with CSOA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

CSOA is a tool for scoring gene set signatures in scRNA-seq data. It constructs, for each input gene, a set of cells highly expressing the gene, then assesses all pairs of cell sets for statistical significance. The significant overlaps are ranked, and the top-ranked overlaps receive scores used to calculate per-cell CSOA scores.

# Installation
To install CSOA, run the following commands in an R session:
```{r setup, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("andrei-stoica26/CSOA")
```

# Prerequisites
In addition to CSOA, you need to install [harmony](https://github.com/immunogenomics/harmony), [patchwork](https://patchwork.data-imaginist.com/index.html),
[scRNAseq](https://bioconductor.org/packages/release/data/experiment/html/scRNAseq.html) and [scuttle](https://www.bioconductor.org/packages/release/bioc/html/scuttle.html) for this tutorial.


# Scoring a gene set

This tutorial uses an scRNA-seq human pancreas dataset with provided cell type annotations. 

After loading the required packages, download the dataset using the `BaronPancreasData` function from `scRNAseq`. The dataset will be stored as a `SingleCellExperiment` object. 

CSOA requires the input scRNA-seq data to be normalized and log-transformed. We will employ the `logNormCounts` function from `scuttle` for this step.
  
```{r message=F, warning=F, results=F}
library(CSOA)
library(ggplot2)
library(harmony)
library(patchwork)
library(scRNAseq)
library(scuttle)
library(Seurat)

sceObj <- BaronPancreasData('human')
sceObj <- logNormCounts(sceObj)
```

We can take a look at the cell types found in this dataset:

```{r}
table(colData(sceObj)$label)
```

Next, we will define a vector of acinar markers based on [PanglaoDB](https://www.panglaodb.se/markers.html):

```{r}
acinarMarkers <- c('PRSS1', 'KLK1', 'CTRC', 'PNLIP', 'AKR1C3', 'CTRB1', 
                   'DUOXA2', 'ALDOB', 'REG3A', 'SERPINA3', 'PRSS3', 'REG1B', 
                   'CFB',	'GDF15',	'MUC1','ANPEP', 'ANGPTL4', 'OLFM4', 
                   'GSTA1', 'LGALS2', 'PDZK1IP1', 'RARRES2', 'CXCL17', 
                   'UBD', 'GSTA2', 'LYZ', 'RBPJL', 'PTF1A', 'CELA3A', 
                   'SPINK1', 'ZG16', 'CEL', 'CELA2A', 'CPB1', 'CELA1', 
                   'PNLIPRP1', 'RNASE1', 'AMY2B', 'CPA2','CPA1', 'CELA3B', 
                   'CTRB2', 'PLA2G1B', 'PRSS2', 'CLPS', 'REG1A', 'SYCN')
```

Before running CSOA, we will convert the `SingleCellExperiment` object to a `Seurat` object in order to visualize the results using `VlnPlot`. Using the input acinar markers, CSOA identifies the acinar cells in the dataset:

```{r warning=F}
seuratObj <- as.Seurat(sceObj)
seuratObj <- runCSOA(seuratObj, acinarMarkers, 'CSOA_acinar')
VlnPlot(seuratObj, 'CSOA_acinar', group.by = 'label') + NoLegend()
```

Alternatively, CSOA can be run on a `SingleCellExperiment` object. The results will be stored as a column in the object's `colData` matrix:

```{r}
sceObj <- runCSOA(sceObj, acinarMarkers, 'CSOA_acinar')
```

Furthermore, CSOA can be run directly on an expression matrix. CSOA requires a dense expression matrix, as opposed to the sparse matrix class (`dgCMatrix`) used by `Seurat` and `SingleCellExperiment`. To satisfy this requirement, the `expMat` function will be employed. This function extracts the log-transformed and normalized expression matrix of the input object—`data` for`Seurat`,  `logcounts` for `SingleCellExperiment`—and converts it to the `matrix` class.

```{r}
geneSetExp <- expMat(seuratObj)
```

When running directly on an expression matrix, CSOA will return a data frame. The CSOA scores are stored in a column named using the `colStr` parameter, set to `'CSOA_acinar'` in our example:

```{r}
df <- runCSOA(geneSetExp, acinarMarkers, 'CSOA_acinar')
head(df)
```

We confirm that the results of the three runs are identical:

```{r}
identical(seuratObj@meta.data$CSOA_acinar, sceObj$CSOA_acinar)
identical(seuratObj@meta.data$CSOA_acinar, df$CSOA_acinar)
```

# Scoring multiple gene sets

CSOA can score multiple gene sets using the `runCSOAMultiple` function. We provide three additional marker lists selected from [PanglaoDB](https://www.panglaodb.se/markers.html):

```{r}
alphaMarkers <- c('GCG', 'TTR', 'PCSK2', 'FXYD5', 'LDB2', 'MAFB',
                  'CHGA', 'SCGB2A1', 'GLS', 'FAP', 'DPP4', 'GPR119',
                  'PAX6', 'NEUROD1', 'LOXL4', 'PLCE1', 'GC', 'KLHL41',
                  'FEV', 'PTGER3', 'RFX6', 'SMARCA1', 'PGR', 'IRX1',
                  'UCP2', 'RGS4', 'KCNK16', 'GLP1R', 'ARX', 'POU3F4',
                  'RESP18', 'PYY', 'SLC38A5', 'TM4SF4', 'CRYBA2', 'SH3GL2', 
                  'PCSK1', 'PRRG2', 'IRX2', 'ALDH1A1','PEMT', 'SMIM24', 
                  'F10', 'SCGN', 'SLC30A8')

ductalMarkers <-	c('CFTR', 'SERPINA5', 'SLPI', 'TFF1', 'CFB', 'LGALS4', 
                   'CTSH',	'PERP', 'PDLIM3',	'WFDC2', 'SLC3A1', 'AQP1', 
                   'ALDH1A3', 'VTCN1',	'KRT19', 'TFF2', 'KRT7', 'CLDN4',
                   'LAMB3', 'TACSTD2', 'CCL2', 'DCDC2','CXCL2', 'CLDN10',
                   'HNF1B', 'KRT20', 'MUC1', 'ONECUT1', 'AMBP', 'HHEX',
                   'ANXA4', 'SPP1', 'PDX1', 'SERPINA3', 'GDF15', 'AKR1C3',
                   'MMP7', 'DEFB1', 'SERPING1', 'TSPAN8', 'CLDN1', 'S100A10',
                   'PIGR')

gammaMarkers <- c('PPY', 'ABCC9', 'FGB', 'ZNF503', 'MEIS1', 'LMO3', 'EGR3', 
                  'CHN2', 'PTGFR', 'ENTPD2', 'AQP3', 'THSD7A', 'CARTPT',
                  'ISL1', 'PAX6', 'NEUROD1', 'APOBEC2', 'SEMA3E', 'SLITRK6',
                  'SERTM1', 'PXK', 'PPY2P', 'ETV1', 'ARX', 'CMTM8', 'SCGB2A1', 
                  'FXYD2', 'SCGN')

```

Now we run CSOA on the four provided marker lists:
```{r warning=F}
geneSets <- list(acinarMarkers, alphaMarkers, ductalMarkers, gammaMarkers)
geneSetNames <- c('CSOA_acinar', 'CSOA_alpha', 'CSOA_ductal', 'CSOA_gamma')
seuratObj <- runCSOAMultiple(seuratObj, geneSets, geneSetNames)
```

To visualize the results of CSOA on multiple gene sets, we define the `patchworkPlot` function, intended to display a plot composed of multiple subplots generated with the same function (specified using the `fun` parameter) as follows:

```{r}
patchworkPlot <- function(seuratObj, geneSetNames, fun, ...){
  plots <- lapply(geneSetNames, function(x) {
  fun(seuratObj, x, ...) + NoLegend() +
    theme(axis.text = element_text(size=10),
          axis.title.x = element_blank(),
          plot.title = element_text (size=11))
    })
  (plots[[1]] | plots[[2]]) / (plots[[3]] | plots[[4]])
}
```

Next, we call the `patchworkPlot` function. For creating each subplot, we use again the `VlnPlot` function from `Seurat`.

```{r warning=F, fig.height=6, fig.width=6}
patchworkPlot(seuratObj, geneSetNames, VlnPlot, group.by = 'label') 
```

Additionally, CSOA can score multiple gene sets using a `SingleCellExperiment` object:

```{r}
sceObj <- runCSOAMultiple(sceObj, geneSets, geneSetNames)
sapply(geneSetNames, 
       function(x) identical(seuratObj@meta.data[[x]], colData(sceObj)[[x]]))
```

CSOA can also score multiple gene sets directly on an expression matrix. The results will again be available as a data frame.

```{r}
df <- runCSOAMultiple(geneSetExp, geneSets, geneSetNames)
head(df)
sapply(geneSetNames, 
       function(x) identical(seuratObj@meta.data[[x]], df[[x]]))
```

# Feature plot visualization

CSOA provides `featureWes`, an adaptation of `FeaturePlot` from `Seurat` with enhanced distinctiveness.

The function builds a continuous scale based on two colors chosen from any [wesanderson](https://github.com/karthik/wesanderson) palette, specified using the following arguments:
-`wesPal`: The name of the palette
`wesLow`: The index of the color chosen for the feature minimum
`wesHigh`: The index of the color chosen for the feature maximum. 

The `idClass` argument can be used to display labels of Seurat identity classes on the plot.

To enable feature plot visualization, we will perform dimensionality reduction on the Seurat object. As the data came from four human subjects, integration is necessary. We will employ `harmony` for this task.

```{r message=F, warning=F, results=F}

seuratObj <- FindVariableFeatures(seuratObj)
seuratObj <- ScaleData(seuratObj)
seuratObj <- RunPCA(seuratObj)
seuratObj <- RunHarmony(seuratObj, group.by.vars = 'orig.ident')
seuratObj <- RunUMAP(seuratObj, dims = 1:20, reduction = 'harmony')

```

Now we can visualize the CSOA results as feature plots:

```{r message=F, warning=F, fig.height=6, fig.width=6}
patchworkPlot(seuratObj, geneSetNames, featureWes, wesPal = 'Darjeeling1', 
              wesLow = 5, wesHigh = 1, idClass = 'label')  
```

# Session information {-}
```{r}
sessionInfo()
```
