---
title: "The CSOA algorithm"
author: "Andrei-Florian Stoica"
package: CSOA
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{The CSOA algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>"
)
```

# Constructing cell sets
First, we repeat the setup from the **Getting started with CSOA** tutorial. We need the `BaronPancreasData` dataset from `scRNAseq` and the list of acinar markers from [PanglaoDB](https://www.panglaodb.se/markers.html).

```{r message=FALSE, warning=FALSE, results=FALSE}
library(CSOA)
library(scRNAseq)
library(scuttle)
library(Seurat)

sceObj <- BaronPancreasData('human')
sceObj <- logNormCounts(sceObj)

seuratObj <- as.Seurat(sceObj)

acinarMarkers <- c('PRSS1', 'KLK1', 'CTRC', 'PNLIP', 'AKR1C3', 'CTRB1', 
                   'DUOXA2', 'ALDOB', 'REG3A', 'SERPINA3', 'PRSS3', 'REG1B', 
                   'CFB',	'GDF15',	'MUC1','ANPEP', 'ANGPTL4', 'OLFM4', 
                   'GSTA1', 'LGALS2', 'PDZK1IP1', 'RARRES2', 'CXCL17', 
                   'UBD', 'GSTA2', 'LYZ', 'RBPJL', 'PTF1A', 'CELA3A', 
                   'SPINK1', 'ZG16', 'CEL', 'CELA2A', 'CPB1', 'CELA1', 
                   'PNLIPRP1', 'RNASE1', 'AMY2B', 'CPA2','CPA1', 'CELA3B', 
                   'CTRB2', 'PLA2G1B', 'PRSS2', 'CLPS', 'REG1A', 'SYCN')
```

We now extract the expression data for the input genes. The results will be stored as a non-sparse matrix:

```{r}
geneSetExp <- expMat(seuratObj, acinarMarkers)
class(geneSetExp)
dim(geneSetExp)
```

Next, a cell set representing the highest-expression cells is constructed for each input gene. The cutoff for cell selection is implemented using the `percentile` argument, set at `90` by default.

**! Note:** The percentile is taken over all the cells expressing the gene, not over all the cells in the dataset. Therefore, the cell sets are allowed to vary in size.

```{r}
cellSets <- percentileSets(geneSetExp, percentile=90)
sapply(cellSets, length)
```

We can now visualize the distribution of the cell sets among all the cells:

```{r, out.height='90%', out.width='90%', fig.height=6, fig.width=5}
mat <- cellDistribution(cellSets, allCells = colnames(seuratObj))
basicHeatmap(mat, c('Gene', 'Cell', 'Presence'), 
             title='Distribution of cell sets among all cells',
             palType='fillDis')
```

The cell sets strongly cluster in a small region of the heatmap, which indicates their significant overlap, expected since they correspond to markers of the same cell type identity.

# Assessing overlaps of pairs of cell sets

The next step is calculating the statistical significance of the overlaps of pairs of cell sets by using hypergeometric tests using the `cellSetsOverlaps` function. The data frame returned by this function can be saved using the `overlapFileName`parameter:

```{r}
cso <- cellSetsOverlaps(cellSets, nCells=dim(seuratObj)[2], 
                        overlapFileName='acinarOverlaps')
```

**! Note:** The `nCells` argument must be set to the total number of cells in the Seurat object, not to the number of cells in the cell sets.

We can inspect the output using `qGrab`, a wrapper around `qs::qread` which deletes a `.qs` file after reading it:


```{r}
df <- qGrab('acinarOverlaps.qs')
dim(df)
head(df)
```

The number of rows (`1081`) equals the number of all possible unordered pairs that can be generated from a list of `47` genes (`47 * 46 / 2 = 1081`).

To perform both steps—creating cell sets and assessing overlaps of pairs of cell sets—in a single call, the `generateOverlaps` function can be used:

```{r}
cso2 <- generateOverlaps(geneSetExp, percentile = 90)
identical(cso, cso2)
```

# Identifying and scoring top overlaps

The output of `cellSetsOverlaps` can be made ready for scoring in a single call, using `processOverlaps`:

```{r}
po <- processOverlaps(cso)
head(po)
```

`processOverlaps` sequentially performs five main steps by calling five helper functions:

1. Pre-filtering: `byCorrectDF`

2. Ranking: `rankOverlaps`

3. Identifying filtering cutoff: `prepareFiltering`

3. Filtering: `filterOverlaps`

4. Scoring: `scoreOverlaps`

`byCorrectDF` performs the Benjamini-Yekutieli correction for multiple testing and filters out insignificant overlaps:

```{r}
po2 <- byCorrectDF(cso)
dim(po2)
head(po2)
```

`rankOverlaps` defines an overlap rank using the following steps:

1. Two preliminary overlap ranks are computed based on: 1) adjusted p-value; 2) recorded-over-expected size ratio.

2. The preliminary ranks are replaced by **connectivity-based ranks** (`pvalRank`, `ratioRank`). For each edge corresponding to an overlap, the p-value rank is replaced by the minimum p-value rank of the vertices neighboring the edge, and the same for the ratio rank.

3. A raw aggregate rank (`rawAggRank`) is computed as the average of `pvalRank` and `ratioRank`.

4. A final aggregate rank (`rank`) defined as the order imposed by the raw aggregate rank is constructed.

```{r}
po2 <- rankOverlaps(po2)
head(po2)
```

This rank is used to select top overlaps. CSOA defines the overlap rank cutoff as the highest-frequency rank, if unique, and as the average of the minimum and maximum of the highest-frequency ranks otherwise. Geometrically, this entails taking the highest point of the upper convex hull of the set of points on a grid with overlap ranks on the `x` axis and frequencies on the `y` axis:

```{r, out.height='100%', out.width='100%', fig.height=5, fig.width=5}
freqDF <- dplyr::count(po2, rank)
overlapCutoffPlot(freqDF, findRankCutoff(freqDF))
```

Next, we run `prepareFiltering`, a wrapper around `findRankCutoff` which returns the highest raw rank of the non-selected overlaps:

```{r, out.height='100%', out.width='100%', fig.height=5, fig.width=6}
firstOutRawRank <- prepareFiltering(po2)
tempDF <- subset(po2, rawAggRank < firstOutRawRank)
v <- c(tempDF$rawAggRank[nrow(tempDF)], firstOutRawRank)
names(v) <- c('lastInRawRank', 'firstOutRawRank')
v
```

We are now ready to run `filterOverlaps`:
```{r}
po2 <- filterOverlaps(po2, firstOutRawRank)
dim(po2)
```

Next, `scoreOverlaps` is applied with the default `log` overlap scoring method, which assigns scores starting at `1` and decreasing logarithmically towards `0` for each unique rank (though `0` is not reached but interpreted as the score of the first non-top overlap). The resulting data frame is identical with the one generated earlier using `processOverlaps`. 

```{r}
po2 <- scoreOverlaps(po2)
identical(po, po2)
unique(po2$rawAggRank)
head(po2$score, 30)
```

Alternatively, overlaps can be scored by performing inverted minmax normalization of the ranks of the top overlaps and `firstOutRawRank`:

```{r}
po2Minmax <- scoreOverlaps(po2, 'minmax', firstOutRawRank)
head(po2Minmax$score, 30)
```

Both methods return positive scores with `1` as the largest value, higher scores for higher-ranked overlaps, and equal scores for equally-ranked overlaps.

# Scoring the cells

Only input genes that registered a top overlap participate in the scoring of cells. We select these genes using `overlapGenes`, and we minmax-normalize their expression:
```{r}
genes <- overlapGenes(po2)
length(genes)
length(acinarMarkers)
normExp <- kerntools::minmax(geneSetExp[genes, ], rows=TRUE)
```

We now compute the per-cell scores of each gene pair corresponding to a top overlap by multiplying these three quantities:

1. The overlap score of the gene pair;

2. The in-cell minmax-normalized expresssion of the first gene;

3. The in-cell minmax-normalized expresssion of the second gene.

```{r}
pcPairScores <- computePCPairScores(po2, normExp)
dim(pcPairScores)
```

The per-cell CSOA score is the minmax-normalized sum of the columns of `pcPairScores`:

```{r}
df <- computePCSetScores(pcPairScores, cellNames = colnames(geneSetExp), 
                         colStr = 'CSOA_acinar')
head(df)
```

Finally, we attach the CSOA scores to the `Seurat` object:

```{r, message=FALSE}
seuratObj <- storeCellScores(seuratObj, df)
head(seuratObj$CSOA_acinar)
```

# The usage of scoring wrappers

CSOA offers several wrappers for scoring the cells. They start and end at different parts of the pipeline, allowing users to explore intermediary outputs and alter intermediary inputs.

The `scoreCells` function requires an unprocessed overlap data frame—the output of `cellSetsOverlaps` or its wrapper `generateOverlaps`—and returns the output of `computePCSetScores`:

```{r}
dfScoreCells <- scoreCells(geneSetExp, cso, 'CSOA_acinar')
identical(dfScoreCells$CSOA_acinar, df$CSOA_acinar)
```

The `computeCellScores` function starts further downstream. It requires a processed overlap data frame (the output of `processOverlaps`), the minmax-normalized expression matrix of top overlap genes, and the cell names. It likewise returns the output of `computePCSetScores`:

```{r}
dfComputeCellScores <- computeCellScores(po2, normExp, 
                                         colnames(geneSetExp), 
                                         'CSOA_acinar')
identical(dfComputeCellScores$CSOA_acinar, df$CSOA_acinar)
```

Finally, the `runCSOA` wrapper runs the whole pipeline including `storeCellScores`:  

```{r}
seuratObj <- runCSOA(seuratObj, acinarMarkers, 'CSOA_acinar')
identical(seuratObj[[]]$CSOA_acinar, df$CSOA_acinar)
```

# Session information {-}
```{r}
sessionInfo()
```

