---
title: "The CSOA algorithm"
author: "Andrei-Florian Stoica"
package: CSOA
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{The CSOA algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Constructing cell sets
First, we repeat the setup from the **Getting started with CSOA** tutorial. We need the `BaronPancreasData` dataset from `scRNAseq` and the list of acinar markers from [PanglaoDB](https://www.panglaodb.se/markers.html).

```{r message=F, warning=F, results=F}
library(CSOA)
library(scRNAseq)
library(scuttle)
library(Seurat)

sceObj <- BaronPancreasData('human')
sceObj <- logNormCounts(sceObj)

seuratObj <- as.Seurat(sceObj)

acinarMarkers <- c('PRSS1', 'KLK1', 'CTRC', 'PNLIP', 'AKR1C3', 'CTRB1', 
                   'DUOXA2', 'ALDOB', 'REG3A', 'SERPINA3', 'PRSS3', 'REG1B', 
                   'CFB',	'GDF15',	'MUC1','ANPEP', 'ANGPTL4', 'OLFM4', 
                   'GSTA1', 'LGALS2', 'PDZK1IP1', 'RARRES2', 'CXCL17', 
                   'UBD', 'GSTA2', 'LYZ', 'RBPJL', 'PTF1A', 'CELA3A', 
                   'SPINK1', 'ZG16', 'CEL', 'CELA2A', 'CPB1', 'CELA1', 
                   'PNLIPRP1', 'RNASE1', 'AMY2B', 'CPA2','CPA1', 'CELA3B', 
                   'CTRB2', 'PLA2G1B', 'PRSS2', 'CLPS', 'REG1A', 'SYCN')
```

We now extract the expression data for the input genes. The results will be stored as a non-sparse matrix:

```{r}
geneSetExp <- expMat(seuratObj, acinarMarkers)
class(geneSetExp)
dim(geneSetExp)
```

Next, a cell set representing the highest-expression cells is constructed for each input gene. The cutoff for cell selection is implemented using the `percentile` argument, set at `90` by default.

**! Note:** The percentile is taken over all the cells expressing the gene, not over all the cells in the dataset. Therefore, the cell sets are allowed to vary in size.

```{r}
cellSets <- percentileSets(geneSetExp, percentile = 90)
sapply(cellSets, length)
```

We can now visualize the distribution of the cell sets among all the cells:

```{r, out.height='100%', out.width='100%'}
mat <- cellDistribution(cellSets, allCells = colnames(seuratObj))
basicHeatmap(mat, c('Gene', 'Cell', 'Presence'), 
             title='Distribution of cell sets among all cells',
             palType='fillDis')
```

The cell sets strongly cluster in a small region of the heatmap, which indicates their significant overlap, expected since they correspond to markers of the same cell type identity.

# Assessing overlaps of pairs of cell sets

The next step is calculating the statistical significance of the overlaps of pairs of cell sets by using hypergeometric tests using the `cellSetsOverlaps` function. The data frame returned by this function can be saved using the `overlapFileName`parameter:

```{r}
cso <- cellSetsOverlaps(cellSets, nCells=dim(seuratObj)[2], 
                        overlapFileName='acinarOverlaps')
```

**! Note:** The `nCells` argument must be set to the total number of cells in the Seurat object, not to the number of cells in the cell sets.

We can inspect the output using `qGrab`, a wrapper around `qs::qread` which deletes a `.qs` file after reading it:


```{r}
df <- qGrab('acinarOverlaps.qs')
dim(df)
head(df)
```

The number of rows (`1081`) equals the number of all possible unordered pairs that can be generated from a list of `47` genes (`47 * 46 / 2 = 1081`).

To perform both steps—creating cell sets and assessing overlaps of pairs of cell sets—in a single call, the `generateOverlaps` function can be used:

```{r}
cso2 <- generateOverlaps(geneSetExp, percentile = 90)
identical(cso, cso2)
```

# Identifying and scoring top overlaps

The output of `cellSetsOverlaps` can be made ready for scoring in a single call, using `processOverlaps`:

```{r}
po <- processOverlaps(cso)
head(po)
```

`processOverlaps` sequentially performs five main steps by calling five helper functions:

1. Pre-filtering: `byCorrectDF`

2. Ranking: `rankOverlaps`

3. Identifying filtering cutoff: `prepareFiltering`

3. Filtering: `filterOverlaps`

4. Scoring: `scoreOverlaps`

`byCorrectDF` first performs the [Benjamini-Yekutieli correction for multiple testing](https://projecteuclid.org/journals/annals-of-statistics/volume-29/issue-4/The-control-of-the-false-discovery-rate-in-multiple-testing/10.1214/aos/1013699998.full). This slightly increases all p-values, but without changing their relative order. Subsequently, `byCorrectDF` filters out insignificant overlaps, with a significance level set by default at `0.05`:

```{r}
po2 <- byCorrectDF(cso)
dim(po2)
head(po2)
```

The overlap data frame has gained a new column (`pvalAdj`). The number of rows has decreased from `1081` to `619` due to the filtering of insignificant overlaps.

Next, `rankOverlaps` is used to define a rank using the following steps:

1. Two preliminary overlap ranks are computed based on: 1) **adjusted p-value**; 2) **recorded-over-expected size ratio**.

2. The preliminary ranks are replaced by **connectivity-based ranks** (`pvalRank`, `ratioRank`). For each edge corresponding to an overlap, the p-value rank is **replaced by the minimum p-value rank of the vertices neighboring the edge**, and the same for the ratio rank.

3. A raw aggregate rank (`rawAggRank`) is computed as the average of `pvalRank` and `ratioRank`.

4. A final aggregate rank (`rank`) defined as the order imposed by the raw aggregate rank is constructed.

```{r}
po2 <- rankOverlaps(po2)
head(po2)
```

We will now select the overlaps that will be retained for score calculation. CSOA defines the overlap rank cutoff as the highest-frequency rank, if unique; if multiple ranks share the maximum frequency, CSOA takes the average of their minimum and maximum. 

Geometrically, this can be interpreted as taking the highest point of the upper convex hull determined by points on a grid with overlap ranks on the `x` axis and frequencies on the `y` axis. We use the `overlapCutoffPlot` function to visualize this procedure:

```{r, out.height='100%', out.width='100%', fig.height=6, fig.width=6}
freqDF <- dplyr::count(po2, rank)
overlapCutoffPlot(freqDF, findRankCutoff(freqDF))
```

We can now run `prepareFiltering`, a wrapper around `findRankCutoff` which returns the highest raw rank of the overlaps that were **not** selected (`firstOutRawRank`). This information can be used later by the `scoreOverlaps` function if the scoring method (`osMethod`) is changed from the default `log` to `minmax`. 

```{r}
firstOutRawRank <- prepareFiltering(po2)
tempDF <- subset(po2, rawAggRank < firstOutRawRank)
v <- c(tempDF$rawAggRank[nrow(tempDF)], firstOutRawRank)
names(v) <- c('lastInRawRank', 'firstOutRawRank')
v
```

We are now ready to run `filterOverlaps`:
```{r}
po2 <- filterOverlaps(po2, firstOutRawRank)
dim(po2)
```

Next, `scoreOverlaps` is applied with the default `log` overlap scoring method. The resulting data frame is identical with the one generated earlier using `processOverlaps`, as default parameters were used in both cases. 

```{r}
po2 <- scoreOverlaps(po2)
head(po2)
identical(po, po2)
```

The `log` method assigns values starting at `1` and decreasing logarithmically towards `0`, corresponding to each unique rank. Overlaps with the same rank receive the same score. The figure of `0` is not reached, but is interpreted as the score of this first non-top overlap. We can plot the scores versus the ranks using `rankScorePlot`:

```{r, out.height='100%', out.width='100%'}
rankScorePlot(po2, 'Overlap scores computed using the log method')
```
The alternative scoring option is `minmax`, which performs inverted minmax normalization of the ranks of the top overlaps and the previously computed `firstOutRawRank`:

```{r, out.height='100%', out.width='100%'}
po2Minmax <- scoreOverlaps(po2, 'minmax', firstOutRawRank)
rankScorePlot(po2Minmax, 'Overlap scores computed using the minmax method')
head(po2Minmax)
```

For both methods, scores are positive, the largest score is `1`, higher-ranked overlaps have higher scores, and equally-ranked overlaps have equal scores.

# Scoring the cells

The `overlapGenes` function selects all genes that registered at least one top overlap. In general, these are **not** the same as the original input genes, because not all input genes register top overlaps. Here, we observe that `27` out of the original `47` input genes have registered top overlaps:
```{r}
genes <- overlapGenes(po2)
length(genes)
length(acinarMarkers)
setdiff(acinarMarkers, genes)
```

We extract and minmax-normalize the expression of these `27` genes:

```{r}
normExp <- kerntools::minmax(geneSetExp[genes, ], rows=T)
```

We can now compute the per-cell scores of each gene pair corresponding to a top overlap by multiplying these three quantities:

1. The overlap score of the gene pair;

2. The in-cell minmax-normalized expresssion of the first gene;

3. The in-cell minmax-normalized expresssion of the second gene.

```{r}
pcPairScores <- computePCPairScores(po2, normExp)
dim(pcPairScores)
```

Next, we compute the CSOA score for each cell by summing the columns of `pcPairScores` and performing minmax normalization to bring the results to `[0, 1]`. We use the `computePCSetScores` function for this task, with the cell names taken from the `geneSetExp` expression matrix. The result is a data frame with a score column:

```{r}
df <- computePCSetScores(pcPairScores, cellNames = colnames(geneSetExp), 
                         colStr = 'CSOA_acinar')
head(df)
```

To attach the scores to the `Seurat` object, we run `storeCellScores`:

```{r, message=F, warning=F, out.height='100%', out.width='100%'}
seuratObj <- storeCellScores(seuratObj, df)
VlnPlot(seuratObj, 'CSOA_acinar', group.by = 'label') + NoLegend()
```

# The usage of scoring wrappers

CSOA offers multiple wrappers for scoring the cells, starting and ending at different parts of the pipeline. These allow users to explore intermediary outputs and alter intermediary inputs.

The `scoreCells` function requires an unprocessed overlap data frame—the output of `cellSetsOverlaps` or its wrapper `generateOverlaps`—and returns the output of `computePCSetScores`:

```{r}
dfScoreCells <- scoreCells(geneSetExp, cso, 'CSOA_acinar')
identical(dfScoreCells$CSOA_acinar, df$CSOA_acinar)
```

The `computeCellScores` function starts further downstream, requiring a processed overlap data frame (the output of `processOverlaps`), the minmax-normalized expression matrix of top overlap genes, as well as the cell names. It likewise returns the output of `computePCSetScores`:

```{r}
dfComputeCellScores <- computeCellScores(po2, normExp, 
                                         colnames(geneSetExp), 
                                         'CSOA_acinar')
identical(dfComputeCellScores$CSOA_acinar, df$CSOA_acinar)
```

Finally, the `runCSOA` wrapper introduced in **Getting started with CSOA** runs the whole pipeline, including `storeCellScores`, the step after `computePCSetScores`:  

```{r}
seuratObj <- runCSOA(seuratObj, acinarMarkers, 'CSOA_acinar')
identical(seuratObj@meta.data$CSOA_acinar, df$CSOA_acinar)
```

# Session information {-}
```{r}
sessionInfo()
```

